---
title: "Drawing polygons with holes in animint2"
author: "Nishita Shah"
date: "2026-08-21"
categories: [gsoc, animint]
---

During [Google Summer of Code 2026](https://github.com/animint/animint2/issues/322) I worked on [animint2](https://github.com/animint/animint2), an R package for animated, interactive grammar-of-graphics visualizations that render in the browser with D3.

One limitation was that `geom_polygon()` could draw filled shapes, but **not polygons with holes**—for example a country with an inland lake, or a region with an inner cutout. That was [issue #252](https://github.com/animint/animint2/issues/252). The fix is now merged in [PR #328](https://github.com/animint/animint2/pull/328).

Hover the plot below. Tooltips appear on the filled blue regions, and stay off inside holes and on the background.

```{r contour-holes}
library(animint2)

m.list <- list(
  island=rbind(
    c(0,0,0,0,0,0,0),
    c(0,1,1,1,1,1,0),
    c(0,1,0,0,0,1,0),
    c(0,1,0,1,0,1,0),
    c(0,1,0,0,0,1,0),
    c(0,1,1,1,1,1,0),
    c(0,0,0,0,0,0,0)),
  hole=rbind(
    c(0,0,0,0,0,0,0),
    c(0,1,1,1,1,1,0),
    c(0,1,0,0,0,1,0),
    c(0,1,0,0,0,1,0),
    c(0,1,0,0,0,1,0),
    c(0,1,1,1,1,1,0),
    c(0,0,0,0,0,0,0)),
  filled=rbind(
    c(0,0,0,0,0,0,0),
    c(0,1,1,1,1,1,0),
    c(0,1,1,1,1,1,0),
    c(0,1,1,1,1,1,0),
    c(0,1,1,1,1,1,0),
    c(0,1,1,1,1,1,0),
    c(0,0,0,0,0,0,0)))

poly.list <- list()
point.list <- list()
for(grp.i in seq_along(m.list)){
  offset <- grp.i * 7
  group <- names(m.list)[[grp.i]]
  m <- m.list[[grp.i]]
  clines <- contourLines(1:ncol(m), 1:nrow(m), m, levels=0.5)
  poly.list[[grp.i]] <- data.table(subgroup=seq_along(clines))[, data.table(
      group,
      x=clines[[subgroup]]$x + offset,
      y=clines[[subgroup]]$y)
  ), by=subgroup]
  point.list[[grp.i]] <- data.frame(
    row=as.integer(row(m)),
    col=as.integer(col(m)) + offset,
    num=as.numeric(m),
    group=group)
}
poly.df <- do.call(rbind, poly.list)
point.df <- do.call(rbind, point.list)

animint(
  contours=ggplot()+
    scale_fill_manual(breaks=0:1)+ # to emphasize that these are the only two data values
    geom_point(aes(
      col, row, fill=num),
      data=point.df,
      color="red", size=3)+
    geom_polygon(aes(
      x, y, group=group, subgroup=subgroup, tooltip=group),
      data=poly.df,
      alpha=0.5,
      fill="steelblue")+
    xlab("Three contour polygons: island (left), hole (middle), filled (right)")+
    theme_animint(width=800, height=300)
)
```

From left to right:

- **island** has a hole with a filled island inside it.
- **hole** is a filled ring with an empty interior.
- **filled** has no hole.

Red points mark the grid. Hover a point inside a hole: there should be no polygon tooltip. Hover a filled cell: the tooltip should show `island`, `hole`, or `filled`.

## Why this is useful

A polygon with a hole is any filled region that should stay **empty on the inside**: land around a lake, a country around an enclave, a building around a courtyard, or a contour ring around a valley.

That is not only a drawing problem. In an interactive plot, the hole has to behave as empty too. If the lake is still treated as part of the country, then:

- the lake is painted with the country's color;
- hovering the lake shows the country's tooltip;
- clicking the lake selects the country.

With holes, the filled land responds to the mouse, and the cutout does not. That is what you want in choropleth maps, spatial data with inland water, and contour-style graphics where some cells are inside a ring and some are not.

The plots in this post are small examples of that idea. Hover a hole: nothing. Hover the blue fill: a tooltip. The same rule applies to clicks and other mouse interactions.

## How to use it

Same idea as ggplot2: `group` identifies each polygon (one filled region), and `subgroup` distinguishes the outer ring from hole rings. The first subgroup is the outer boundary; later subgroups are holes. Several separate polygons each get their own `group`. If you leave `subgroup` off, inner rings are filled as well.

You do not need a special geom. Use `geom_polygon()` as usual, add `subgroup`, and tooltips or selection on that layer will already skip the holes.

The square below is the same idea with no contour machinery: outer ring filled, inner ring empty. Hover the blue fill versus the hole.

```{r square-hole}
square.hole <- data.frame(
  x=c(0, 4, 4, 0,  1.2, 2.8, 2.8, 1.2),
  y=c(0, 0, 4, 4,  1.2, 1.2, 2.8, 2.8),
  subgroup=c(rep("outer", 4), rep("inner", 4)),
  region="square")

animint(
  square=ggplot()+
    geom_polygon(aes(
      x, y, group=region, subgroup=subgroup,
      tooltip="filled region (not the hole)"),
      data=square.hole,
      fill="steelblue",
      alpha=0.7)+
    coord_equal()+
    ggtitle("A square with a square hole")+
    theme_animint(width=360, height=360)
)
```

## The problem

In ggplot2, polygons with holes use the `subgroup` aesthetic: one outer ring and one or more inner rings. animint2 did not fully support that path for interactive SVG output.

That gap has a timeline. The original `animint` package (2013) depended on ggplot2. Incompatible ggplot2 changes in 2014–2017 led to the **2018** `animint2` fork, which copies the relevant ggplot2 code and no longer tracks ggplot2 releases. ggplot2 then added `geom_polygon(aes(subgroup=...))` in [**3.2.0 (June 2019)**](https://www.tidyverse.org/blog/2019/06/ggplot2-3-2-0/), after R 3.6 gained `grid::pathGrob()` support for holes. Because that landed after the fork, animint2's interactive renderer never picked it up until [issue #252](https://github.com/animint/animint2/issues/252) / [PR #328](https://github.com/animint/animint2/pull/328) in 2026.

Without holes in the SVG output:

- Inner regions were still treated as filled.
- Hit-testing and tooltips could appear **inside** areas that should be empty.
- Maps and contour-style polygons looked wrong in the interactive plot.

We needed the outer ring filled and the holes empty, including for mouse interaction and tooltips.

## How it works

From the user's point of view, a hole is just an inner boundary that should stay empty. The renderer treats that inner region as not part of the polygon: it is not painted, and the mouse does not hit it. That is why a tooltip can appear on land and disappear over a lake, without any extra click-handling code on your side.

Under the hood the compiler notices when polygon data have a `subgroup` column, and the browser draws those rings with even-odd fill so holes stay empty for both painting and hit-testing.

## The approach

The fix spans **R** (data and metadata) and **JavaScript** (SVG rendering).

### 1. R: recognize `subgroup` for polygons

In `geom_polygon`, `subgroup` was added to the default aesthetics. When polygon data actually contains a `subgroup` column, the compiler sets a flag and stores subgroup as character data so the JS side knows holes are present:

```r
if(g$geom == "polygon" && "subgroup" %in% names(g.data)){
  g$data_has_subgroup <- TRUE
  g.data$subgroup <- as.character(g.data$subgroup)
  g$types[["subgroup"]] <- "character"
}
```

### 2. JavaScript: GeoJSON + even-odd fill

For normal polygons we still build a simple SVG path. For polygons with holes we:

1. Nest points by `subgroup` (outer ring + hole rings).
2. Build a GeoJSON `Polygon` (closed rings).
3. Render with `d3.geo.path()`.
4. Set SVG `fill-rule: evenodd` so holes stay unfilled.

```javascript
if(g_info.geom === "polygon" && g_info.data_has_subgroup){
  var geoPath = d3.geo.path().projection(null);
  eActions = function(e){
    e.attr("d", function(d){
      var points = keyed_data[d.value];
      var nested = d3.nest()
        .key(function(pt){ return pt.subgroup; })
        .entries(points);
      var coords = nested.map(function(group){
        var ring = group.values.map(function(pt){
          return [scales.x(pt.x), scales.y(pt.y)];
        });
        if(ring.length > 0){
          ring = ring.concat([ring[0]]);
        }
        return ring;
      });
      var geojson = {
        type: "Polygon",
        coordinates: coords
      };
      return geoPath(geojson);
    })
    .style("fill-rule", "evenodd");
  };
}
```

## Challenges along the way

This PR taught me as much about open-source process as about SVG.

1. **Matching the requested JS style.** An early path-string approach was replaced with `d3.geo.path()`, matching the preferred design from related work and review feedback.

2. **Keeping the diff focused.** Porting between branches created a noisy Files changed history. Cleaning that so the PR only showed the real polygon-hole work was important for review.

3. **CI vs real bugs.** After merge conflicts and DESCRIPTION fixes (author-list syntax, blank lines that broke `pak`), the main test suite went green. Remaining atime noise was a known infrastructure issue ([#340](https://github.com/animint/animint2/issues/340)), not a hole-rendering bug.

4. **Credit and git email.** The PR was opened by Toby as a strong failing test on master; my implementation landed on that branch. Contributor credit is recorded in `NEWS` and `DESCRIPTION`. Separately, some commits used a mistyped email, so GitHub did not always link them to my profile—something I will keep consistent going forward.

## What I learned

- Interactive graphics need **data (R)**, **rendering (JS)**, and **behavior tests** to stay correct together.
- SVG details like `fill-rule: evenodd` are not cosmetic—they define hit-testing.
- Good mentor tests that fail before a fix (and pass after) make review much clearer.
- Small packaging mistakes (`DESCRIPTION` commas / blank lines) can block an otherwise correct feature on CI.

## Closing

I am grateful that [PR #328](https://github.com/animint/animint2/pull/328) is merged and that animint2 users can now draw polygons with holes via `subgroup`, with tooltips that respect those holes.

Thanks to [Toby Dylan Hocking](https://github.com/tdhock) and the animint community for the reviews, tests, and patience through the polish cycle.

**Links**

- PR: <https://github.com/animint/animint2/pull/328>
- Issue: <https://github.com/animint/animint2/issues/252>
- GSoC updates: <https://github.com/animint/animint2/issues/322>
- Package: <https://animint.github.io/animint2/>
